\documentclass[a4paper,UKenglish]{oasics-v2016}
%This is a template for producing OASIcs articles. 
%See oasics-v2016-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
% for section-numbered lemmas etc., use "numberwithinsect"
 
\usepackage{microtype}%if unwanted, comment out or use option "draft"

\usepackage[normalem]{ulem}

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the recommended bibstyle

\newcommand{\edit}[1]{{\color{red} #1 }}

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{KubeNow: a cloud agnostic platform for microservice-oriented applications\footnote{This work was supported by the PhenoMeNal H2020 consortium.}}
%\titlerunning{A Sample OASIcs Article} %optional, in case that the title is too long; the running title should fit into the top page column

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
%\author[1]{John Q. Open}
%\author[2]{Joan R. Access}
%\affil[1]{Dummy University Computing Laboratory, Address/City, Country\\
%  \texttt{open@dummyuniversity.org}}
%\affil[2]{Department of Informatics, Dummy College, Address/City, Country\\
%  \texttt{access@dummycollege.org}}
%\authorrunning{J.\,Q. Open and J.\,R. Access} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\author[1,2]{Marco Capuccini}
\author[3]{Anders Larsson}
\author[1]{Salman Toor}
\author[2]{Ola Spjuth}
%\author[1]{Ola Spjuth}
\affil[1]{Department of Information Technology, Uppsala University\\
  \texttt{marco.capuccini@it.uu.se}\\
  \texttt{salman.toor@it.uu.se}}
\affil[2]{Department of Pharmaceutical Biosciences, Uppsala University\\
  \texttt{marco.capuccini@farmbio.uu.se}\\
  \texttt{ola.spjuth@farmbio.uu.se}}
\affil[3]{Department of Cell and Molecular Biology, Uppsala University\\
  \texttt{anders.larsson@icm.uu.se}}
  
\authorrunning{M. Capuccini and O. Spjuth} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{M. Capuccini and O. Spjuth}%mandatory, please use full first names. OASIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\subjclass{D.2.11 Software Architectures}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
\keywords{Microservices; Cloud computing; Infrastructure as Code; Docker; Kubernetes}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Fergus Leahy and Juliana Franco}
\EventNoEds{2}
\EventLongTitle{2017 Imperial College Computing Student Workshop (ICCSW 2017)}
\EventShortTitle{ICCSW 2017}
\EventAcronym{ICCSW}
\EventYear{2017}
\EventDate{September 26—-27, 2017}
\EventLocation{London, UK}
\EventLogo{../graphics/iccsw17}
\SeriesVolume{60}
\ArticleNo{9}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


% Page numbers in side margins
\usepackage[some]{background}
\usepackage{ifthen}
\SetBgContents{\textcolor{gray}{\LARGE \bf \hl \thepage}}
\SetBgOpacity{1}
\SetBgScale{1}
\SetBgAngle{0}
\SetBgColor{black}
\makeatletter
  \AddEverypageHook{%
    \ifthenelse{\isodd{\thepage}}%
      {\SetBgPosition{.9\paperwidth,-.9\paperheight}%
	  \thispagestyle{empty}%
	}%
      {\SetBgPosition{.1\paperwidth,-.9\paperheight}%
	  \thispagestyle{empty}%
	}%
    \bg@material}
\makeatother
\begin{document}

\maketitle

\begin{abstract}

KubeNow is a platform for rapid and continuous deployment of microservice-based applications over cloud infrastructure. Within the field of software engineering, the microservice-based architecture is a methodology in which complex applications are divided into smaller, more narrow services. These services are independently deployable and compatible with each other like building blocks. These blocks can be combined in multiple ways, according to specific use cases. Microservices are designed around a few concepts: they offer a minimal and complete set of features, they are portable and platform independent, they are accessible through language agnostic APIs and they are encouraged to use standard data formats. These characteristics promote separation of concerns, isolation and interoperability, while coupling nicely with test-driven development. Among many others, some well-known companies that build their software around microservices are: Google, Amazon, PayPal Holdings Inc. and Netflix \cite{biomicro}. 

% Cutting on this, to give more weight to the work we have done
%Software containers and language agnostic APIs, which constitute the enabling underlying technology for microservices, have been available for decades \cite{jails,soap,rest}. Nevertheless, the process of building and orchestrating microservices over computer clusters, or even within a single machine, introduces a layer of complexity that is difficult to cope with. Software frameworks that ease this process appeared in the open source ecosystem only recently, giving place to a massive and quick adoption in the information technology industry. Among many others, some well-known companies that build their software around microservices are: Google, Amazon, PayPal Holdings Inc. and Netflix \cite{biomicro}. 

Cloud computing is a new technology trend that enables the allocation of virtual infrastructure on demand, giving place to a new business model where organizations can purchase resources with a pay-per-use pricing arrangement \cite{clouds}. Microservices in cloud environments can help to build scalable and resilient applications, with the goal of maximing resource usage and reducing costs. At the time of writing, Docker and Kubernetes are the most broadly adopted container engine and container orchestration framework \cite{docker,kubernetes}. Even though these software tools ease microcroservices operations considerably, their setup and configuration is still complex, tedious and time consuming. When allocating cloud resources on demand this becomes a critical issue, since applications need to be continuously deployed and scaled, possibly over different cloud providers, to minimize infrastructure costs. This new challenging way of provisioning infrastructure was the main motivation for the development of KubeNow.

% Cutting on this due to 2 pages limit
%Some commercial and open source products to automate Docker/Kubernetes deployment on cloud exist \cite{tectonic,gce,kops,kargo}, but their adoption is usually cost prohibitive and/or tightly coupled to a few cloud providers. Furthermore, none of the available open source deployment tools covers the application layer deployment, which can be nontrivial and time consuming. 

KubeNow provides the means to rapidly deploy fully configured clusters, automating Docker and Kubernetes configuration, while providing a mechanism for the application layer setup. We designed KubeNow using the Infrastructure as Code (IaC) paradigm, meaning that the virtual resources and the provisioning process are defined as machine-readable language. A natural consequence of this choice is that KubeNow is immutable and repeatable over different cloud providers, being cloud agnostic in this sense. In addition, IaC enables infrastructure version control and collaborative development. 

KubeNow has been adopted by the PhenoMeNal H2020 consortium as the platform used to launch on demand Cloud Research Environments (CRE) \cite{phenomenal}. The PhenoMeNal CRE allows for running reproducible large-scale medical metabolomics analysis. In addition, we are currently developing additional software layers for large-scale analysis on top of KubeNow including: Apache Spark \cite{spark}, Pachyderm \cite{pachyderm} and Slurm \cite{slurm}. KubeNow supports Amazon Web Services \cite{aws}, Google Compute Engine \cite{gcloud} and OpenStack \cite{openstack}. The software is generally applicable and publicly available as open source on GitHub \cite{kubenow}.

 \end{abstract}
\end{document}
