Welcome to the 2015 Imperial College Computing Student Workshop (ICCSW’15), the
fifth workshop in its series. ICCSW was initiated with a “by students, for
students” spirit: a workshop organised solely by students to give student
speakers the opportunity to present their work. The organising students gain
the valuable experience of what is involved in conference organisation,
including writing a call for sponsors, taking part in the reviewing process,
and chairing a session. On the other hand, the participating students benefit
from the interaction with international researchers who are at a similar stage
in their career and developing skills in presenting their research to a
non-specialist computer science audience. This year’s Imperial College
Computing Student Workshop was organised by the Imperial College ACM Student
Chapter closely following the “by students, for students” spirit. As in
previous years, a unique reviewing model was adopted, where in addition to
members of the Imperial College ACM Student Chapter all authors were involved
in the reviewing process. Furthermore, we reached out to students at a variety
of European universities to act as “ambassadors” for ICCSW’15 by spreading the
word about the workshop.

This volume contains the papers accepted for presentation at the 2015 Imperial
College Computing Student Workshop. ICCSW’15 received 18 submissions from 10
different countries across two different tracks: the traditional paper track
including technical and tool papers, and a survey track. After the thorough
reviewing process and discussion by members of the Imperial College ACM Student
Chapter 11 papers were accepted, representing a 61\% acceptance rate which is in
line with previous years.

In addition to the regular presentations of accepted papers, we are pleased to
welcome two prominent keynote speakers to ICCSW’15:

\begin{itemize}
  \item ``Google and Open Source'' by Chris DiBona, Director of Open Source at
Google
  \item ``Going Beyond Fact-Based Question Answering'' by Erik Mueller,
researcher at MIT and member of the IBM team that developed the Watson
Jeopardy! system.
\end{itemize}

On behalf of the organising committee, we wish to thank all authors, accepted
or not, as well as our ambassadors, who acted as reviewers in our unique peer
review process and promoted ICCSW’15 respectively. Furthermore, we also wish to
thank our sponsors: Google, for their platinum-level sponsorship, who have
supported ICCSW since its inception in 2011; and ARM, for their bronze-level
sponsorship.  Without the support of our sponsors, ICCSW’15 would not have been
possible. Our thanks also goes to Imperial College London, and in particular
the Department of Computing, who provided us with invaluable support in
organising ICCSW’15.
\vspace{5ex}

\noindent Claudia Schulz and Daniel Liew \\
\textbf{ICCSW’15 Editors}
